---
title: "Using EDITO Datalab: A Complete Guide for Marine Researchers"
subtitle: "From RStudio to Jupyter to VSCode - Everything you need to know"
author: "EDITO Tutorial Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Introduction

Welcome to the comprehensive guide for using EDITO Datalab! This tutorial is designed for marine researchers who want to leverage cloud computing and modern data formats for their research.

## What is EDITO Datalab?

**EDITO** (European Digital Twin of the Ocean) provides a powerful cloud platform for marine research. The Datalab offers:

- **RStudio Service**: For R-based analysis and visualization
- **Jupyter Service**: For Python-based machine learning and data science
- **VSCode Service**: For larger projects and multi-language development
- **Personal Storage**: Secure, persistent storage for your data
- **ARCO Data Access**: Analysis Ready Cloud Optimized data formats

## Why This Matters for Marine Research

- **Large datasets**: Handle millions of marine records efficiently
- **Environmental data**: Access oceanographic data for analysis
- **Collaboration**: Share code and data with research teams
- **Reproducibility**: Version control and documented workflows
- **Scalability**: Process data that won't fit on your local computer

---

# Getting Started with RStudio Service

## Launching RStudio

1. Go to [EDITO Datalab](https://datalab.dive.edito.eu/)
2. Select "RStudio" from the service catalog
3. Configure resources (CPU, memory) as needed
4. Launch the service

## Basic Data Access

```{r load-packages}
# Load required packages
library(arrow)      # For reading Parquet files (ARCO data)
library(sf)         # For spatial data handling
library(dplyr)      # For data manipulation
library(ggplot2)    # For plotting
library(rstac)      # For accessing EDITO STAC API
```

```{r explore-stac}
# Connect to EDITO STAC API
stac_endpoint <- "https://api.dive.edito.eu/data/"
collections <- stac(stac_endpoint) %>%
  rstac::collections() %>%
  get_request()

# See what data collections are available
cat("Available data collections:\n")
for(i in 1:min(10, length(collections$collections))) {
  cat(i, ":", collections$collections[[i]]$id, "\n")
}
```

## Working with Biodiversity Data

```{r biodiversity-data}
# Read EUROBIS occurrence data (contains marine biodiversity observations)
parquet_url <- "https://s3.waw3-1.cloudferro.com/emodnet/biology/eurobis_occurrence_data/eurobis_occurrences_geoparquet_2024-10-01.parquet"

# Read a sample of the data
biodiversity_occurrences <- arrow::read_parquet(parquet_url) %>%
  head(1000)

# Filter for marine species (including fish, invertebrates, algae, etc.)
marine_data <- biodiversity_occurrences %>%
  filter(grepl("fish|Fish|pisces|Pisces|mollusca|Mollusca|algae|Algae|crustacea|Crustacea", scientificName, ignore.case = TRUE) |
         grepl("fish|Fish|mollusk|Mollusk|algae|Algae|crab|Crab", vernacularName, ignore.case = TRUE))

cat("Found", nrow(marine_data), "marine biodiversity occurrence records\n")
```

```{r biodiversity-visualization}
# Create a simple map of marine biodiversity occurrences
if(nrow(marine_data) > 0) {
  # Convert to spatial data
  marine_sf <- st_as_sf(marine_data, wkt = "geometry")
  
  # Plot
  ggplot(marine_sf) +
    geom_sf(aes(color = scientificName), size = 0.5) +
    labs(title = "Marine Biodiversity Occurrences from EDITO Data",
         subtitle = paste("Total records:", nrow(marine_sf)),
         color = "Species") +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

---

# Using Jupyter Service for Python Analysis

## Launching Jupyter

1. Go to [EDITO Datalab](https://datalab.dive.edito.eu/)
2. Select "Jupyter" from the service catalog
3. Choose Python environment
4. Launch the service

## Python Data Analysis

```python
# Example Python code for Jupyter
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns

# Create sample marine biodiversity data
np.random.seed(42)
marine_data = pd.DataFrame({
    'species_id': np.repeat(range(1, 51), 20),
    'timestamp': pd.date_range('2024-01-01', '2024-12-31', periods=1000),
    'latitude': np.random.uniform(50, 60, 1000),
    'longitude': np.random.uniform(0, 10, 1000),
    'depth': np.random.uniform(5, 200, 1000),
    'temperature': np.random.normal(10, 3, 1000)
})

# Basic analysis
print(f"Dataset shape: {marine_data.shape}")
print(f"Species count: {marine_data['species_id'].nunique()}")
print(f"Date range: {marine_data['timestamp'].min()} to {marine_data['timestamp'].max()}")
```

## Machine Learning for Habitat Modeling

```python
# Simple habitat suitability modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.model_selection import train_test_split

# Create habitat suitability classes
marine_data['habitat_suitability'] = (
    (marine_data['temperature'] > 8) & 
    (marine_data['temperature'] < 15) &
    (marine_data['depth'] > 20) &
    (marine_data['depth'] < 150)
).astype(int)

# Prepare features
features = ['latitude', 'longitude', 'depth', 'temperature']
X = marine_data[features]
y = marine_data['habitat_suitability']

# Split and train
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)
model = RandomForestClassifier(n_estimators=100, random_state=42)
model.fit(X_train, y_train)

# Evaluate
accuracy = model.score(X_test, y_test)
print(f"Model accuracy: {accuracy:.3f}")
```

---

# VSCode for Larger Projects

## When to Use VSCode

VSCode is perfect for:
- Multi-language projects (R + Python)
- Large codebases with many files
- Git integration and version control
- Collaborative development
- Complex data processing pipelines

## Project Organization

```
marine_research_project/
├── data/
│   ├── raw/                 # Original marine data
│   ├── processed/           # Cleaned data
│   └── environmental/       # Environmental data
├── scripts/
│   ├── data_processing/     # Data cleaning
│   ├── analysis/           # Statistical analysis
│   └── visualization/      # Plotting
├── notebooks/
│   ├── exploration.ipynb   # Data exploration
│   └── analysis.ipynb      # Main analysis
└── outputs/
    ├── figures/            # Generated plots
    └── reports/            # Generated reports
```

---

# Personal Storage Integration

## Accessing Your Personal Storage

Your personal storage credentials are automatically available in EDITO services:

```{r personal-storage}
# Check if storage credentials are available
if(Sys.getenv("AWS_ACCESS_KEY_ID") != "") {
  cat("✅ Personal storage credentials found!\n")
  cat("Storage endpoint:", Sys.getenv("AWS_S3_ENDPOINT"), "\n")
  cat("Region:", Sys.getenv("AWS_DEFAULT_REGION"), "\n")
} else {
  cat("❌ No storage credentials found. Make sure you're running in EDITO Datalab.\n")
}
```

## Saving Data to Personal Storage

```{r save-data}
# Example: Save processed marine data
if(nrow(marine_data) > 0) {
  # Save as CSV
  write.csv(marine_data, "marine_analysis_results.csv", row.names = FALSE)
  
  # Save as Parquet (more efficient)
  arrow::write_parquet(marine_data, "marine_analysis_results.parquet")
  
  cat("✅ Data saved locally\n")
  cat("- marine_analysis_results.csv\n")
  cat("- marine_analysis_results.parquet\n")
}
```

## Python Storage Access

```python
# Python example for personal storage
import boto3
import os

# Connect to EDITO's MinIO storage
s3_client = boto3.client(
    's3',
    endpoint_url=f"https://{os.getenv('AWS_S3_ENDPOINT')}",
    aws_access_key_id=os.getenv('AWS_ACCESS_KEY_ID'),
    aws_secret_access_key=os.getenv('AWS_SECRET_ACCESS_KEY'),
    aws_session_token=os.getenv('AWS_SESSION_TOKEN'),
    region_name=os.getenv('AWS_DEFAULT_REGION')
)

# Save data to personal storage
s3_client.put_object(
    Bucket='your-bucket-name',
    Key='marine_research/analysis_results.csv',
    Body=marine_data.to_csv(index=False),
    ContentType='text/csv'
)
```

---

# Working with ARCO Data

## What is ARCO Data?

**ARCO** = Analysis Ready Cloud Optimized

- **Parquet**: Columnar format, perfect for tabular data
- **Zarr**: Chunked arrays, ideal for raster data
- **COG**: Cloud Optimized GeoTIFF for imagery

## Reading Parquet Data

```{r parquet-example}
# Parquet is much faster than CSV for large datasets
# Example: Reading marine biodiversity occurrence data
parquet_url <- "https://s3.waw3-1.cloudferro.com/emodnet/biology/eurobis_occurrence_data/eurobis_occurrences_geoparquet_2024-10-01.parquet"

# Read only specific columns (memory efficient)
biodiversity_sample <- arrow::read_parquet(parquet_url) %>%
  select(scientificName, vernacularName, decimalLatitude, decimalLongitude, eventDate) %>%
  head(1000)

cat("Loaded", nrow(biodiversity_sample), "records\n")
print(head(biodiversity_sample))
```

## Python Parquet Access

```python
# Python example for Parquet data
import pyarrow.parquet as pq
import pandas as pd

# Read Parquet data
parquet_url = "https://s3.waw3-1.cloudferro.com/emodnet/biology/eurobis_occurrence_data/eurobis_occurrences_geoparquet_2024-10-01.parquet"

# Read with specific columns
columns = ['scientificName', 'decimalLatitude', 'decimalLongitude', 'eventDate']
df = pd.read_parquet(parquet_url, columns=columns)

print(f"Loaded {len(df)} records")
print(df.head())
```

---

# Best Practices for Marine Research

## Data Organization

1. **Use consistent naming conventions**
2. **Organize data in logical folders**
3. **Include metadata files**
4. **Document your data processing steps**

## Code Management

1. **Use version control (Git)**
2. **Write clear comments**
3. **Test your code regularly**
4. **Save intermediate results**

## Collaboration

1. **Share code via GitHub/GitLab**
2. **Use personal storage for data sharing**
3. **Document your methods**
4. **Make your work reproducible**

---

# Troubleshooting Common Issues

## RStudio Issues

- **Package installation**: Use `install.packages()` in the console
- **Memory issues**: Increase memory allocation in service configuration
- **Data loading**: Use `arrow::read_parquet()` for large files

## Jupyter Issues

- **Kernel problems**: Restart the kernel
- **Package installation**: Use `!pip install package_name` in cells
- **Memory issues**: Process data in chunks

## VSCode Issues

- **Terminal access**: Use Ctrl+` to open terminal
- **Git integration**: Use the Source Control panel
- **Extensions**: Install from the Extensions marketplace

---

# Next Steps

## Explore More

1. **Try different services**: RStudio, Jupyter, VSCode
2. **Access more data**: Explore EDITO's data collections
3. **Learn about data sharing**: Use personal storage effectively
4. **Join the community**: Connect with other researchers

## Resources

- [EDITO Datalab](https://datalab.dive.edito.eu/)
- [Personal Storage](https://datalab.dive.edito.eu/account/storage)
- [EDITO Tutorials](https://dive.edito.eu/training)
- [Data API Documentation](https://pub.pages.mercator-ocean.fr/edito-infra/edito-tutorials-content/#/interactWithTheDataAPI)

---

# Conclusion

EDITO Datalab provides powerful tools for marine research:

- **RStudio**: Perfect for statistical analysis and visualization
- **Jupyter**: Ideal for machine learning and data exploration
- **VSCode**: Great for larger, multi-language projects
- **Personal Storage**: Secure, persistent data management
- **ARCO Data**: Fast, efficient access to large datasets

Start with the service that matches your current workflow, then explore others as your needs grow. The cloud-based approach means you can work with datasets that would be impossible to handle on your local computer.

Happy analyzing! 🌊🐠

---

*This tutorial was created for the EDITO Datalab workshop. For questions or support, contact edito-infra-dev@mercator-ocean.eu*
